Robert C. Dunnell’s _Systematics in Prehistory_ at 50

### 2022-02-15 analysis/paper/paper.Rmd
edits from co-authors

---
title: "Robert C. Dunnell’s _Systematics in Prehistory_ at 50"
author:
  - Felix Riede:
      email: f.riede@cas.au.dk
      institute: [UofO]
      correspondence: true
  - Astolfo Araujo:
      institute: [UofSP]
      correspondence: false
  - Ben Marwick:
      email: bmarwick@uw.edu
      institute: [UofWA]
      correspondence: false
institute:
  - UofO: Department of Archaeology and Heritage Studies, Aarhus University, Moesgård Allé 20, 8270 Højbjerg, Denmark
  - UofSP: Museum of Archaeology and Ethnology, University of São Paulo, Brazil
  - UofWA: Department of Anthropology, University of Washington, Seattle, WA, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/evolutionary-anthropology.csl" # Insert path for the bib-style
abstract: |
  2021 marks the fiftieth anniversary of the publication of Robert C. Dunnell’s (1971) diminutive yet dense _Systematics in Prehistory_. At the height of the debate between Culture History and New Archaeology, Dunnell’s work sought to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Paleolithic: systematics. Dunnell himself was notorious and controversial, however, but the importance of his work remains underappreciated. Like other precocious works of that tumultuous time Systematics in Prehistory today remains absent from most course reading lists and gathers dust on library shelves. In this contribution we argue for a greater appreciation of its as yet unfulfilled conceptual and analytical promise. In particular, we briefly chart its somewhat delayed impact via evolutionary archaeology, including how it has also influenced non-Anglophone traditions, especially in South America. The obstinate persistence of classification issues in paleoanthropology and Paleolithic archaeology, we argue, warrants a second look at Dunnell's _Systematics_.
---

@@

# but this breaks kniting with the knit button in RStudio
# only way to knit is to run targets::tar_make() in the R 
# console

@@

The notion of major paradigm shifts in the history of archaeology is characterized by a great deal of hyperbole and rhetorical maneuvering. That said, the late 1960s and 1970s really do stand out as rather revolutionary. In the US, the Binfordian juggernaut increasingly hammered away at its culture-historical nemesis, while in the UK, David Clarke's formidable *Analytical Archaeology* was shaking up the establishment with novel concepts and methods as well as incisive rhetoric [@clarke1968a]. The reception and impact of these works has seen a great deal of attention in later years, especially as many researchers are rediscovering the merits of Clarke's approach in particular [@nicholas2012; @lycett2018a]. The year 2021 marks the fiftieth anniversary of the publication of Robert C. Dunnell's [-@dunnell1971] diminutive yet dense *Systematics in Prehistory*, a volume that is concerned with nothing but classification in archaeology. Published at the height of the debate between Culture History and New Archaeology, Dunnell's work was no less iconoclastic than that of Binford or Clarke, but it was considerably narrower in its endeavor to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Paleolithic: systematics.

Dunnell himself was notorious and controversial, and his *Systematics* was received with mixed reactions by his contemporaries who commented - mostly negatively - both on his style of writing and the book's content [@spaulding1974; @tuggle1974; @bayard1973; @shenkel1973]. Like other precocious works of that tumultuous time *Systematics in Prehistory* today remains absent from most course reading lists and gathers dust on library shelves. As Lyman has recently shown, systematics takes up little space in contemporary archaeological research or teaching [@lyman2021]. If at least some of the success of Clarke and Binford can be attributed to them boldly tackling exciting and large-scale topics such as migration and adaptation using then avant-garde terminology, then the corresponding obscurity of Dunnell's _Systematics_ can perhaps be attributed to him focusing on an issue that simply seemed too tedious  Furthermore, archaeology has moved ahead in such a way that leaves some of the key claims _Systematics_ exposed as a distinct outliers in modern American archaeology. For example, his insistence that the exclusive focus of archaeology be the physical traces of pasts human activity, and that archaeology cannot be both a science and a sub-discipline of anthropology, are generally minority positions now, with archaeology continuing to be a sub-field of anthropology in most US universities. In countries where prehistory is situated in different institutional contexts, this particular issue is not a major concern. Be it as it may, Dunnell was correct in his axiomatic insistence that rigorous classification comes before any other analysis or interpretation. Without consistent and explicit classification, any scientific discipline will inevitably fail to produce cumulative insights.

@@

For Dunnell himself, the soul-searching that began with writing _Systematics in Prehistory_ led him to discover evolutionary theory. In a series of follow-up papers, he forcefully argued for the benefits of a scientific and Darwinian archaeology [@dunnell1980; @dunnell1982]. While his writing style did not make him many friends, these arguments have since become foundational for the development of evolutionary archaeology, especially in the Americas [@obrien1996]. Initially, following the direct lead of Dunnell, this approach was rather narrowly selectionist - treating artefacts as the hard parts of the human phenotype and selection acting on these as the main driver of change [@obrien1990] - but since has become more fully aligned with cultural evolutionary thinking in the form of dual-inheritance and niche construction theory [@riede2019; @marwick2006]. Vitally, cultural evolutionary theory and its focus on the transmission of cultural knowledge via various modes of learning has provided the crucial generative mechanism for material culture systematics. In its contemporary form, selection, but also drift and a range of transmission biases, play important roles in explaining culture change.

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the idiosyncrasies of archaeological classification are nowhere more apparent and acute than in the archaeology of human evolutionary history. Rooted in French antiquarianism, the development of Paleolithic systematics has been likened to 'accidents of history' [@clark2009], and there are few periods or regions of the Paleolithic that have not seen debate about the validity or otherwise of their analytical units [@reynolds2019]. The use of older typological classifications remains prevalent, despite clear and repeated critiques [e.g. @bisson2000]. More recently, the analysis of technological traits has supplemented or even eclipsed purely typological approaches. However, theoretical explications of the generative mechanisms, and rigorous comparative systematics backed by transparent and replicable analytics as demanded by Dunnell, remain exceptions rather than the rule [@tostevin2013]. Today, a great deal of attention is again being paid to systematics in prehistory, and as Barton and Clark [-@barton2021] have pointed out, the continuing adherence to outmoded classifications is preventing the exploration of more relevant and pressing research questions. Several researchers are tackling classificatory issues with novel and mostly quantitative means [@ivanovait2020; @grove2021; @leplongeon2020]. At the same time, however, the topic remains poorly heeded in archaeological pedagogy [@lyman2021]. Given that cultural evolutionary theory itself teaches us that aspects of culture most easily and rapidly change when scaffolded through active teaching [@Riede_Walsh_Nowell_Langley_Johannsen_2021], we recommend that, after half a century, Dunnell's _Systematics in Prehistory_ - and with it rigorous and replicable ways of classifying material culture - are placed more abundantly on our curricula, in addition to continuing the ongoing critique and transformation of existing classifications through novel research. Only when the construction and meaning of our analytical units and their relationships among one another are transparent, and robust cultural taxonomies are in place can we seriously hope to understand the patterns and processes that have shaped cultural evolution in deep history.

## 3 THE CONTRIBUTION OF _SYSTEMATICS_

To survey the contribution that *Systematics in Prehistory* has made to evolutionary anthropology, we take inspiration from its concept of statistical clustering as a method for organising variability, and recent developments in the statistical analysis of text. We searched for items citing _Systematics in Prehistory_ on Google Scholar, downloaded all the search result pages (data were collected on July 2021), and extracted bibliographic data from each work citing *Systematics*. The raw data and R code for our analysis of the Google Scholar results are openly available online at http://doi.org/10.17605/OSF.IO/JBPFW.

@@

We found `r how_many_citing_works` citations of _Systematics_; for reference the most highly cited archaeology publication from the same year is Binford's [-@binford1971mortuary] 'Mortuary Practices: Their Study and Their Potential' with 1621 citations currently on Google Scholar. Citations to _Systematics_ have steadily accumulated over time, with a distinct increase in the annual rate of citation in the early 2000s as evolutionary concepts become increasingly integrated into archaeological science (Figure \@ref(fig:sip-citations-of-citing-work)). Several works that cite _Systematics_ have themselves been very highly cited, with ten publications receiving over 500 citations, demonstrating its influence over a range of topics including social theory [@Shanks_Tilley_1987], behavioral archaeology [@schiffer2016behavioral], and lithics [@Andrefsky_1998]. Looking at the top fifty words in the titles of the works citing _Systematics_, we can also see that 'lithics' and 'stone' are prominent, indicating that its contribution was especially noted by archaeologists working on stone artefacts (Figure \@ref(fig:sip-keywords-in-titles)). Pottery is the only other artefact type where analysts substantially engaged with _Systematics_. Notable in the title keywords are concepts relating to evolution, for example, 'transmission', 'variation', and 'evolutionary'. Evolution is a very minor theme in _Systematics_, but these word frequency data show it has proven to be a foundational text in applications of evolutionary theory to explaining variability in the archaeological record. Beyond the Anglosphere, we see citations from a small number works in Russian and European languages, but the impact of _Systematics_ is most evident in Spanish-language publications by scholars working on South American stone artefact assemblages.

@@

These South American studies have been especially innovative, taking to heart the critique in _Systematics_ that archaeological typologies are often intuitive, arbitrary, and difficult to replicate by other researchers. Motivated by this critique, scholars such as Marcelo Cardillo, Judith Charlin [-@Cardillo_Charlin_2018] and others have conducted pioneering work in the application of geometric morphometry to stone artefact assemblages to provide a materialist view of technological variation where the focus is on continuous quantitative phenomena. While geometric morphometry has been applied by archaeologists to a range of regions and artefact types (including ceramic and metal), what makes this South American work remarkable as part of the legacy of *Systematics* is their exploration of phylogenetic comparative methods to model and quantify technological variation and change over space and time [e.g. @cardillo2015evolution]. These works show that one of the most important contributions of _Systematics_ has been as a bridge between archaeology and cognate fields, for example, palaeontology, dealing with challenges of classification and concerned with modelling macroevolutionary processes. The bridge-building that has followed from this has been very fruitful, reflected, for example, in the increasing number of archaeological presentations at meetings of the Cultural Evolution Society.

@@

The South American research stands out as a distinctly coherent and exclusive topic (Topic 59) in our topic model of titles of works citing _Systematics_ (Figure \@ref(fig:sip-topic-model)). A topic model is an unsupervised classification of a collection of documents, in this case titles, that uses a probabilistic model to generate mixtures of words that represent themes or topics in the collection [-@Roberts_Stewart_Tingley_others_2014; @Roberts_Stewart_Airoldi_2016]. This finds natural groups of topics similar to how clustering on numeric data finds groups of similar items. We generated a topic model using Latent Dirichlet Allocation method, which resulted in an optimal number of `r how_many_topics` topics. The topic model shows the persistence of the core themes of _Systematics_ in the citing literature, with the most abundant and central topic (Topic 53) about the measurement of material culture variation, and transmission of that variation. The topic model provides additional insights into the contribution of _Systematics_ that are not evident in the word frequency analysis. The archaeology of the central and southern Pacific Ocean stands out, includes works on Polynesian fish hooks, fabrics and stone adzes. Rapa Nui has its own topic, representing the work of Carl Lipo and colleagues on pottery, monuments and population dynamics of that island and the region. Lipo's work is thoroughly grounded in a concern for rigorous and explicit unit construction and innovative testing of archaeological systematics, noting that '[s]ystematics enables us to move beyond common sense' [@lipoisland2021]. We see a concern for the construction of analytical units in areas as diverse as the Palaeolithic of Africa, ceramics of the Maya and Papua New Guinea, and several regions of the Unites States.

While the influence of *Systematics* has extended well beyond Dunnell's own primary study area of the US Southeast, there are some notable limits to its contributions. The Portuguese translation has received only 71 citations, and the Spanish translation just 30. In our data we also see only a single work in an East or Southeast Asian language, a book chapter in Japanese surveying the literature on cultural phylogenetics, indicating limited attention to _Systematics_ from researchers in the Eastern hemisphere. Could cultural differences in reasoning styles [@henrich2010weirdest] result in a diminished relevance for the philosophical content of _Systematics_ outside of Western academic communities? Or perhaps it is simply that the book is quite difficult to read, as noted by reviewers when it was first published ('confusion followed by grudging agreement' is how Spaulding [-@spaulding1974 p. 515-516] described his reaction to the book), and acknowledged by Dunnell in his forword to the 2002 edition. The low readability of the book likely limited the accessibility of its contents for readers whose first language is not English.

A second limitation, or maybe better described as an unrealized contribution, of _Systematics_ is the absence of a distinct research area and community of archaeological systematics. While biology has a Society of Systematic Biologists, and attendant journals such as _Systematic Biology_, _Systematics and Evolution_, and _Systematics and Biodiversity_, archaeology has no equivalent focal points for discussions of systematics that transcends specific geographical and chronological concerns. _Systematics_ was not a work that served as an engine for moving archaeologists together at scale to tackle questions about the formation of units of measurement and classification. Instead these discussions typically happen deep within discrete and disconnected archaeological research communities, constrained by the culture and norms of those groups.

Dunnell intended _Systematics_ to establish a distinctive form of scientific archaeology, and it faced overwhelming competition from the program advocated by New Archaeology. The approaches of the New Archaeologists dominated the literature in the decade after the publication of _Systematics_, thoroughly eclipsing it as a discipline-defining text. Nevertheless, what makes _Systematics_ a classic contribution is that is provides archaeologists with the master key to release themselves from the 'prison of de Mortillet' [@shea2016stone p. xvii], namely the inherited and entrenched analytical habits of research traditions dominated by prestige economies (Gabriel de Mortillet, 1821–1898, was an archaeologist who published the first widely used classification of the Paleolithic, much of which remains in use today). _Systematics_ endures as a striking and precocious provocation to archaeologists to be transparent, rigorous, precise, and deliberate about how we divide up and aggregate material culture and the measurements we take from it to describe and explain the human experience in the past. With archaeologists increasingly pursuing ambitious questions about evolutionary processes that require large scale syntheses of disparate datasets [cf. @Perreault_2019], the message of _Systematics_ will only become more relevant.

@@

>>>>

---
title: "Robert C. Dunnell’s _Systematics in Prehistory_ at 50"
author:
  - Felix Riede:
      email: f.riede@cas.au.dk
      institute: [UofO]
      correspondence: true
  - Astolfo Araujo:
      institute: [UofSP]
      correspondence: false
  - Ben Marwick:
      email: bmarwick@uw.edu
      institute: [UofWA]
      correspondence: false
institute:
  - UofO: Department of Archaeology and Heritage Studies, Aarhus University, Moesgård Allé 20, 8270 Højbjerg, Denmark
  - UofSP: Museum of Archaeology and Ethnology, University of São Paulo, Brazil
  - UofWA: Department of Anthropology, University of Washington, Seattle, WA, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: false 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/evolutionary-anthropology.csl" # Insert path for the bib-sty
abstract: |
  2021 marks the fiftieth anniversary of the publication of Robert C. Dunnell’s (1971) diminutive yet dense _Systematics in Prehistory_. At the height of the debate between Culture History and New Archaeology, Dunnell’s work sought to address a more fundamental issue that was and still is relevant to all branches of prehistoric archaeology, and especially to the study of the Paleolithic: systematics. Dunnell himself was notorious and controversial, however, but the importance of his work remains underappreciated. Like other precocious works of that tumultuous time Systematics in Prehistory today remains absent from most course reading lists and gathers dust on library shelves. In this contribution we argue for a greater appreciation of its as yet unfulfilled conceptual and analytical promise. In particular, we briefly chart its somewhat delayed impact via evolutionary archaeology, including how it has also influenced non-Anglophone traditions, especially in South America. The obstinate persistence of classification issues in paleoanthropology and Paleolithic archaeology, we argue, warrants a second look at Dunnell's _Systematics_.
---

@@

# but this breaks kniting with the knit button in RStudio
# only way to knit is to run targets::tar_make() in the R 
# console. We do not use the knit button at all.

@@

Evolutionary Archaeology is concerned with the study of cultural evolution through material culture proxies and commonly over extended timescales not readily accessible by other means. While evolutionary thinking *per se* is not new in the discipline, an operational evolutionary archaeology emerged only recently, and on the back of several major paradigm transitions [@prentiss2019handbook\]. The notion of major paradigm shifts in the history of archaeology is, however, characterized by a great deal of hyperbole and rhetorical maneuvering. That said, the late 1960s and 1970s do stand out as revolutionary. In the US, the Binfordian juggernaut increasingly hammered away at its culture-historical nemesis, while in the UK, David Clarke's [-@clarke1968a] formidable *Analytical Archaeology* was shaking up the establishment with novel concepts and methods as well as incisive rhetoric. The reception and impact of these works has seen a great deal of attention in later years, especially as many researchers working within ecological and evolutionary approaches to past culture change are rediscovering the merits of Clarke's conceptual approach in particular, and Binford's extensive data synthesis [@nicholas2012; @lycett2018a]. The year 2021 marked the fiftieth anniversary of the publication of Robert C. Dunnell's [-@dunnell1971] diminutive yet dense *Systematics in Prehistory*, a volume that is concerned entirely with classification in archaeology. Appearing at the height of the debate between Culture History and New Archaeology, Dunnell's work was no less iconoclastic than that of Binford or Clarke, but it was considerably narrower in its goal of addressing a more fundamental issue that was and still is relevant to all branches of archaeology: systematics.

Dunnell himself was notorious and controversial, and his *Systematics* was received with mixed reactions by his contemporaries who commented - mostly negatively - both on his style of writing and the book's content [@spaulding1974; @tuggle1974; @bayard1973; @shenkel1973]. Like other precocious works of that tumultuous time *Systematics in Prehistory* today remains absent from most course reading lists and gathers dust on library shelves. As Lyman has recently shown, systematics takes up little space in contemporary archaeological research or teaching [@lyman2021]. In striking contrast, biological systematics is a well developed field with faculty positions, courses, and journals dedicated to this fundamental scientific concern. If at least some of the success of Clarke and Binford can be attributed to them boldly tackling exciting and large-scale topics such as migration and adaptation adaptation through the introduction of avant-garde terminology, then the corresponding obscurity of Dunnell's *Systematics* can perhaps be attributed to him focusing on an issue that simply seemed too quotidian. Furthermore, archaeology has moved ahead in such a way that leaves some of the key claims *Systematics* exposed as a distinct outliers in modern American archaeology. For example, his insistence that the exclusive focus of archaeology be the physical traces of pasts human activity, and that archaeology cannot be both a science and a sub-discipline of anthropology, are generally minority positions now, with archaeology continuing to be a sub-field of anthropology in most US universities at least. In countries where archaeology is situated in different institutional contexts, this particular issue is not a major concern. Be it as it may, Dunnell was correct in his axiomatic insistence that rigorous classification comes before any other analysis or interpretation. Without consistent and explicit classification, any scientific discipline will inevitably fail to produce cumulative insights; certainly, evolutionary analysis would hardly be possible without robust classification.

@@

For Dunnell himself, the soul-searching that began with writing *Systematics in Prehistory* led him to discover evolutionary theory. In a series of follow-up papers, he forcefully argued for the benefits of a scientific and Darwinian archaeology [@dunnell1980; @dunnell1982]. While his writing style did not attract many followers, these arguments have since become foundational for the development of evolutionary archaeology, especially in the Americas [@obrien1996]. Initially, following the direct lead of Dunnell, this approach was rather narrowly selectionist, treating artefacts as the hard parts of the human phenotype and selection acting on these as the main driver of change [@obrien1990]. Evolutionary archaeology has since become both more plural and more fully aligned with cultural evolutionary thinking in the form of behavioural ecology, dual-inheritance and niche construction theory [@riede2019; @marwick2006; @prentiss2021theoretical]. Vitally, cultural evolutionary theory and its focus on the transmission of cultural knowledge via various modes of learning has provided the crucial generative mechanism for material culture systematics. In its contemporary form, selection, but also drift and a range of transmission biases, play important roles in explaining culture change.

While Dunnell worked exclusively on the Holocene prehistory of the Americas, the idiosyncrasies of archaeological classification are nowhere more apparent and acute than in the archaeology of human evolutionary deep history: the Palaeolithic. Rooted in French antiquarianism, the development of Paleolithic systematics, for instance, has been likened to so many 'accidents of history' [@clark2009], but there are few periods or regions of the Palaeolithic that have not seen debate about the validity or otherwise of their analytical units [@reynolds2019]. The use of older typological classifications remains prevalent, despite clear and repeated critiques [e.g. @bisson2000]. More recently, the analysis of technological traits has supplemented or even eclipsed purely typological approaches. However, theoretical explications of the generative mechanisms, and rigorous comparative systematics backed by transparent and replicable analytics as demanded by Dunnell, remain exceptions rather than the rule [@tostevin2013]. Today, a great deal of attention is again being paid to systematics in prehistory, and as Barton and Clark [-@barton2021] have pointed out, the continuing adherence to outmoded classifications is preventing the exploration of more relevant and pressing research questions. Several researchers are tackling classificatory issues with novel and mostly quantitative means [@ivanovait2020; @grove2021; @leplongeon2020]. At the same time, however, the topic remains poorly heeded in archaeological pedagogy [@lyman2021]. Given that cultural evolutionary theory itself teaches us that aspects of culture most easily and rapidly change when scaffolded through active teaching [@Riede_Walsh_Nowell_Langley_Johannsen_2021], we recommend that, after half a century, Dunnell's *Systematics in Prehistory* - and with it rigorous and replicable ways of classifying material culture - are placed more abundantly on our curricula, in addition to continuing the ongoing critique and transformation of existing classifications through novel research. Only when the construction and meaning of our analytical units and their relationships among one another are transparent, and robust cultural taxonomies are in place can we seriously hope to understand the patterns and processes that have shaped cultural evolution in deep history.

## 3 THE CONTRIBUTION OF *SYSTEMATICS*

To survey the contribution that *Systematics in Prehistory* has made to evolutionary anthropology, we take inspiration from its concept of statistical clustering as a method for organising variability, and recent developments in the statistical analysis of text. We searched for items citing *Systematics in Prehistory* on Google Scholar, downloaded all the search result pages (data were collected on July 2021), and extracted bibliographic data from each work citing *Systematics*. The raw data and R code for our analysis of the Google Scholar results, and the R Markdown source document for this paper are openly available online at <http://doi.org/10.17605/OSF.IO/JBPFW>.

@@

We found `r how_many_citing_works` citations of *Systematics*; for reference the most highly cited archaeology publication from the same year is Binford's [-@binford1971mortuary] 'Mortuary Practices: Their Study and Their Potential' with 1621 citations currently on Google Scholar at the time of our data collection. Citations to *Systematics* have steadily accumulated over time, with a distinct increase in the annual rate of citation in the early 2000s as evolutionary concepts become increasingly integrated into archaeological science (Figure \@ref(fig:sip-citations-of-citing-work)). Several works that cite *Systematics* have themselves been very highly cited, with ten publications receiving over 500 citations, demonstrating its influence over a range of topics including social theory [@Shanks_Tilley_1987], behavioral archaeology [@schiffer2016behavioral], and lithic analysis [@Andrefsky_1998]. Looking at the top fifty words in the titles of the works citing *Systematics*, we can also see that 'lithics' and 'stone' are prominent, indicating that its contribution was especially noted by archaeologists working on stone artefacts (Figure \@ref(fig:sip-keywords-in-titles)). Pottery is the only other artefact type where analysts substantially engaged with *Systematics*. Notable in the title keywords are concepts relating to evolution, for example, 'transmission', 'variation', and 'evolutionary'. Evolution is a very minor theme in *Systematics* itself, but these word frequency data show it has proven to be a foundational text in applications of evolutionary theory to explaining variability in the archaeological record.

@@

Beyond the Anglosphere, we see citations from a small number works in Russian and European languages (Figure \@ref(fig:sip-keywords-in-titles)), but the impact of *Systematics* is most evident in Spanish-language publications by scholars working on South American stone artefact assemblages. These South American studies have been especially innovative, taking to heart the critique in *Systematics* that archaeological typologies are often intuitive, arbitrary, and difficult to replicate by other researchers. Motivated by this critique, scholars such as Marcelo Cardillo, Judith Charlin [-@Cardillo_Charlin_2018\] and others have conducted pioneering work in the application of geometric morphometrics to stone artefact assemblages in an effort to provide a materialist view of technological variation where the focus is on continuous quantitative phenomena. While geometric morphometrics has been applied by archaeologists to a range of regions and artefact types (including ceramic and metal), what makes this South American work remarkable as part of the legacy of *Systematics* is their exploration of modern phylogenetic comparative methods to model and quantify technological variation and change over space and time [e.g. @cardillo2015evolution]. These works show that one of the most important contributions of *Systematics* has been as a bridge between archaeology and cognate fields - first and foremost palaeontology - dealing with challenges of classification and concerned with modelling macroevolutionary processes. The bridge-building that has followed from this has been very fruitful, reflected, for example, in the increasing number of archaeological presentations at meetings of the Cultural Evolution Society since the first one in 2017.

@@

The South American research stands out as a distinctly coherent and exclusive topic (Topic 59) in our topic model of titles of works citing *Systematics* (Figure \@ref(fig:sip-topic-model)). A topic model is an unsupervised classification of a collection of documents, in this case titles of works citing *Systematics*, that uses a probabilistic approach to generate mixtures of words that represent themes or topics in the collection [-@Roberts_Stewart_Tingley_others_2014; @Roberts_Stewart_Airoldi_2016]. This finds natural groups of topics similar to how clustering on numeric data finds groups of similar items. We generated a topic model with the widely-used Latent Dirichlet Allocation method, which resulted in an optimal number of `r how_many_topics` topics. The topic model shows the persistence of the core themes of *Systematics* in the citing literature, with the most abundant and central topic (Topic 53) about the measurement of material culture variation, and transmission of that variation. The topic model provides additional insights into the contribution of *Systematics* that are not evident in the word frequency analysis. For example, the archaeology of the central and southern Pacific Ocean stands out, and includes works on Polynesian fish hooks, fabrics and stone adzes. Rapa Nui has its own topic, representing the work of Carl Lipo and colleagues on pottery, monuments and population dynamics of that island and the region. Lipo's work is thoroughly grounded in a concern for rigorous and explicit unit construction and innovative testing of archaeological systematics, noting that 'systematics enables us to move beyond common sense' [@lipoisland2021]. We see a concern for the construction of analytical units in areas as diverse as the Palaeolithic of Africa, ceramics of the Maya and Papua New Guinea, and several regions of the Unites States.

The influence of *Systematics* has extended well beyond Dunnell's own primary study area of the North American Southeast. For example, the Portuguese translation published in 2007 has received 75 citations, and the Spanish translation 30. At face value these numbers seem low, but when we take into consideration that Binford's Portuguese translation of *In Pursuit of the Past* published in 1991 has 31 citations, we can have a better perspective of the influence Dunnell's work. On the other hand, in our data we see only a single work in an East or Southeast Asian language, a book chapter in Japanese surveying the literature on cultural phylogenetics, indicating limited attention to *Systematics* from researchers in the Eastern hemisphere. Could cultural differences in reasoning styles [@henrich2010weirdest] result in a diminished relevance for the philosophical content of *Systematics* outside of Western academic communities? Or perhaps it is simply that the book is quite difficult to read, as noted by reviewers when it was first published ('confusion followed by grudging agreement' is how Spaulding [-@spaulding1974 p. 515-516] described his reaction to the book), and acknowledged by Dunnell in his foreword to the 2002 edition. The low readability of the book has likely limited the accessibility of its contents for readers whose first language is not English, and made it inherently less attractive as a
pedagogical resource.

A second limitation of *Systematics* is that it failed to motivate the formation of a distinct research area and community of archaeological systematics. While biology has a Society of Systematic Biologists, and attendant journals such as *Systematic Biology*, *Systematics and Evolution*, and *Systematics and Biodiversity*, archaeology currently has no equivalent focal points for discussions of systematics that transcends specific geographical and chronological concerns. *Systematics* was not a work that served as an engine for moving archaeologists together at scale to tackle questions about the formation of units of measurement and classification. Instead, these discussions typically happen deep within discrete and disconnected archaeological research communities, constrained by the culture and norms of those groups. The result of this fragmentation is a proliferation of bespoke classifications that lack applicability across cases. It also inhibits reproducibility, synthesis, and the transfer of ideas and innovations across research communities, which may limit the sustainability of archaeology as a discipline.

Dunnell intended *Systematics* to establish a distinctive form of scientific archaeology, and it faced overwhelming competition from the program advocated by the New Archaeology on the one hand, while also meeting resistance from deeply entrenched traditional approaches to classification. In addition, the volume also appeared just before personal computers were beginning to have a major impact on archaeological research practice [@clarkQuantificationAmericanArchaeology1982; @aldenderfer1987assessing\]. The rigor advocated by Dunnell would have lent itself smoothly to the sorts of computational approaches now increasingly adopted in the discipline. Yet, the approaches of the New Archaeologists dominated the literature in the decade after the publication of *Systematics*, thoroughly eclipsing it as a discipline-defining text. Nevertheless, what makes *Systematics* a classic contribution is that is provides archaeologists with the master key to release themselves from the 'prison of de Mortillet' [@shea2016stone p. xvii], namely the inherited and entrenched analytical habits of research traditions dominated by economies of personal prestige (Gabriel de Mortillet, 1821--1898, was an archaeologist who published the first widely used classification of the Paleolithic, much of which remains in use today). *Systematics* endures as a striking and precocious provocation to archaeologists to be transparent, rigorous, precise, and deliberate about how we divide up and aggregate material culture and the measurements we take from it to describe and explain the human experience in the past. With archaeologists increasingly pursuing ambitious questions about evolutionary processes that require large scale syntheses of disparate datasets [cf. @Perreault_2019], the message of *Systematics* will only become more relevant.

@@

This work is part of the CLIOARCH project, an ERC Consolidator Grant project that has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No. 817564).

# Author biographies

Felix Riede is Professor of Prehistoric Archaeology at Aarhus University and PI in the ERC-funded project CLIOARCH (<https://cas.au.dk/en/clioarch/>) that is revisiting Paleolithic cultural taxonomies and aims to put these on more robust epistemological and empirical foundations. His main areas of research are the early prehistory of Northern Europe, pioneer dispersals, cultural phylogenetics and human cognitive evolution.

Astolfo Araujo is Professor at the Museu de Arqueologia e Etnologia da Universidade de São Paulo. He is a leading South American scholar in the application of evolutionary thinking in archaeology. He has researched and written extensively on the cultural taxonomy of prehistoric Brazil and on the history of archaeology in the region.

Ben Marwick is Professor at the Department of Anthropology, University of Washington -- Robert C. Dunnell's former department. He is widely known for his pioneering work in quantitative methods and reproducibility in archaeology, and his applications of cultural evolutionary and behavioral ecological models to the prehistory of mainland Southeast Asia and Australia in particular.

# **Data Availability Statement**

The raw data and R code for our analysis of the Google Scholar results are openly available online at <http://doi.org/10.17605/OSF.IO/JBPFW>.
